# Reproducible Research - Project 1

```{r}
library(tidyverse)
```

**1. Loading data **  
```{r, cahe = TRUE}
data_raw <- read_csv("activity.csv")
```


**2. Plotting a histogram of daily number of steps**
```{r}


daily <- data_raw %>%
  group_by(date) %>%
  summarise(daily_steps = sum(steps, na.rm = TRUE))

ggplot(daily) + geom_histogram(aes(daily_steps), fill = "darkred", binwidth = 1000) + 
  ggtitle("Distribution of daily number of steps") + xlab("Daily steps")
```


**3. Calculating mean and median of daily number of steps**  
Mean value of daily steps equaled to **`r round(mean(daily$daily_steps, na.rm = TRUE))`** and median equaled to **`r median(daily$daily_steps, na.rm = TRUE)`**.

**4. Plotting time series of average steps taken per given 5 minute interval**
```{r}
ts <- data_raw %>%
  group_by(interval) %>%
  summarise(mean_steps = mean(steps, na.rm = TRUE))

ggplot(ts) + geom_line(aes(interval, mean_steps, group = 1), col = "darkred", lwd = 1) +
  ggtitle("Average number of steps made per subsequent 5-minute intervals") + xlab("Interval") + 
  ylab("Mean number of steps") 
```

**5. Indicating a 5-minute interval with maximum average number of steps**
```{r}
max_steps <- filter(ts, mean_steps == max(mean_steps))
```
The interval with the maximum average number of daily steps was interval **`r max_steps$interval`**.

**6. Replacing NAs with 5-minute interval mean**  
The number of NAs in this dataset is `r sum(is.na(data_raw))`. They are replaced with mean number of steps per 5-minute interval.
``` {r}
data_imp <- data_raw %>%
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))
```

**7. Plotting a histogram of daily number of steps after imputation**  
```{r}
daily_imp <- data_imp %>%
  group_by(date) %>%
  summarise(daily_steps = sum(steps))

ggplot(daily_imp) + geom_histogram(aes(daily_steps), fill = "darkred", binwidth = 1000) + 
  ggtitle("Distribution of daily number of steps after imputation") + xlab("Daily steps")  
```  

After the imputation the mean value equaled to **`r round(mean(daily_imp$daily_steps))`** and the median was **`r median(daily_imp$daily_steps)`**. Due to imputation the values are higher.

**8. Inspecting distributions of activity during weekdays and weekends**
```{r}
act_comparison <- data_imp %>%
  mutate(day_type = ifelse(weekdays(date) %in% c("sobota", "niedziela"), "weekend", "weekday")) %>%
  group_by(day_type, interval) %>%
  summarise(mean_steps = mean(steps, na.rm = TRUE))

ggplot(act_comparison) + geom_line(aes(interval, mean_steps, group = 1), col = "darkred", lwd = 1) +
  facet_wrap(~day_type, ncol = 1) + ggtitle("Activity during weekdays and weekends - comparison") + xlab("Interval") + 
  ylab("Mean number of steps") 
```