---
output:
  pdf_document: default
  html_document: default
---
# The socio-economic effects of weather events in United States

## Synopsis
This analysis concerns weather events in USA occuring between 1950 and 2011 and theirs socio-economic effects. It was deduced that most of these events had taken place in Texas, mostly in form of hails and thunderstorm winds. The amount of the events rose substantially in recent years, nevertheless the quantity increased because of more complete data. By far the most harmful events were tornados which had influence on lives of around 100 000 people. In terms of economic effects - Hurricane Catrina which influenced the condition of buildings badly and floods & blizzards in the 20ties.  

## Data Processing
The data used in this article comes from the U.S. National Oceanic and Atmospheric Administration's storm database.The description of the data can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

Some support packages are readed.
```{r, message = FALSE}
library(tidyverse)
library(stringr)
library(lubridate)
```

### Reading in the data
Firstly the raw data obtained from the bz2 archive is loaded. The fields are automatically parsed by the **readr** package and the file is loaded with headers.
```{r, cache = TRUE, message = FALSE}
data_raw <- read_csv("storm.csv.bz2")
```
### Preprocessing
Some data processing was performed. Date was changed to appropriate format and desired columns were selected for further analysis (also changed columns names).
```{r}
data <- data_raw %>%
  select(State = STATE, Date = BGN_DATE, County = COUNTYNAME, Event = EVTYPE,
         Fatalities = FATALITIES, Injuries = INJURIES, Property_Damage = PROPDMG, Crop_Damage = CROPDMG,
         Property_Damage_Scale = PROPDMGEXP, Crop_Damage_Scale = CROPDMGEXP) %>%
  mutate(Date = mdy(str_extract(Date, "\\d{1,2}/\\d{1,2}/\\d{4}")),
         Year = year(Date))
```

## EDA - quick overview
Here is a summary of number of events per state. As can be seen, during the total period Texas was most frequently afflicted. There were around 50k events also in Kansas and Oklahoma.
```{r}
data %>%
  count(State) %>%
  arrange(desc(n))
```
When it comes to the distribution of events - the following table shows these which where noted most often. 
```{r}
data %>%
  count(Event) %>%
  arrange(desc(n)) 
```
USA was damaged mainly but hails, however it seems that there are several codenames for the same event - thunderstorm wind. After some recoding one can discover that those were mostly observed - more than 300k times!
```{r}
data <- data %>%
  mutate(Event = ifelse(Event %in% c("TSTM WIND", "THUNDERSTORM WINDS"), "THUNDERSTORM WIND", Event))

data %>%  
  count(Event) %>%
  arrange(desc(n))
```
Finally - the following plot shows the number of events during the whole period. It looks like the overall quantity rose substantially. Nevertheless, it has to be noted that in the earlier years there were simply not many good records noted. Beware of drawing too hasty conclusions!
```{r}
data %>%
  count(Year) %>%
  arrange(Year) %>%
ggplot() + geom_line(aes(Year, n), col = "darkred", lwd = 1) + ggtitle("Yearly number of events") +
  ylab("")
```

## Results
There were two questions addressed by this analysis:  
  **1. Which type of events were most harmful with respect to population health?**  
  **2. Which types of events had the greatest economic consequences?**
  
The following chart explains which events were most harmful for the population as a whole, meaning the sum of injured and fatalities. Although thunderstorm tornados were much more rare they had the biggest influence - there were almost 100k citizens affacted by this event overall. The rest of the phenomenas concerned around 10k Americans.
```{r}
victims <- data %>%
  group_by(Event) %>%
  summarise(Fatalities = sum(Fatalities),
            Injuries = sum(Injuries)) %>%
  mutate(Victims = Fatalities + Injuries) %>%
  arrange(desc(Victims)) %>%
  slice(1:5)

ggplot(victims) + geom_col(aes(reorder(Event, -Victims), Victims), fill = "darkred") +
  ggtitle("The most harmful events for the population") + xlab("Event")
```
When it comes to economic impact - below are three most costly events in the history in terms of property damage and crop damage measured in Billions.
```{r}
data %>%
  filter(Property_Damage_Scale == "B") %>%
  arrange(desc(Property_Damage)) %>%
  slice(1:3) %>%
  select(State, Date, Event, Property_Damage)

data %>%
  filter(Crop_Damage_Scale == "B") %>%
  arrange(desc(Crop_Damage)) %>%
  slice(1:3) %>%
  select(State, Date, Event, Crop_Damage)
```
The first value concerns floods in California in 2006. However, it's most likely a typing mistake and the value should have been stated in millions - a quick glance at the Wikipedia shows, that the damages cost overall around $300 millions. Positions 2nd and 3rd are more probable - these were the financial effects after hurricane Catrina which occured in Lousiana - it was the costliest weather phenomenon in USA history.  

In terms of crop damages - there were 2 major events in the history. First one was *Great Flood of 1993* which took place in Illinois. Second concerned ice sttorms in Missisipi in 1994. Both events cost around $5 billions.